---
title: "An overview of package plot3logit"
author: "Flavio Santi, Maria Michela Dickson, Giuseppe Espa"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An overview of package plot3logit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.asp = 1
)
```




## Introduction

The package permits the covariate effects of trinomial regression models to be
represented graphically by means of a ternary plot. The aim of the plots is
helping the interpretation of regression coefficients in terms of the effects
that a change in regressors' values has on the probability distribution of the
dependent variable. Such changes may involve either a single regressor, or a
group of them (composite changes), and the package permits both cases to be
handled in a user-friendly way. Theoretical and methodological details are
illustrated and discussed in Santi, Dickson and Espa (2019).

The package can read the results of **both categorical and ordinal trinomial
logit** regression fitted by various functions (see the next section) and
creates a `field3logit` object which may be represented by means of functions
`gg3logit` and `stat_3logit`.

The `plot3logit` package inherits graphical classes and methods from the package
`ggtern` (Hamilton 2018) which, in turn, is based on the package `ggplot2`
(Wickham 2016).

Graphical representation based on **standard graphics** is made available
through the package `Ternary` (Smith 2017) by functions `plot3logit` and
`TernaryField`, and by the `plot` method of `field3logit` objects.

See the help of `field3logit` for representing composite effects and
`multifield3logit` for drawing multiple fields.



## Compatibility

Function `field3logit` of package `plot3logit` can read trinomial regression
estimates from the output of the following functions:

* `multinom` of package `nnet` (logit regression);
* `polr` of package `MASS` (ordinal logit regression);
* `mlogit` of package `mlogit` (logit regression).

Moreover, explicit matrix of regression coefficients can be passed to
`field3logit`. See the help of the package (type `? 'plot3logit-package'`) for
furhter details.



## A brief example


Fit a trilogit model by means of package `nnet` where the student's employment
situation is analysed with respect to all variables in the dataset
`cross_1year`:
```{r, message = FALSE}
library(plot3logit)
data(cross_1year)
library(nnet)
mod0 <- multinom(employment_sit ~ ., data = cross_1year)
```

The gender effect is analysed by means of a ternary plot which is generated in
two steps, however, package `plot3logit` should be loaded:
```{r}
library(plot3logit)
```

Firstly, the vector field is computed:
```{r}
field0 <- field3logit(mod0, 'genderFemale')
```

Secondly, the field is represented on a ternary plot, using either
`gg`-graphics:
```{r}
gg3logit(field0) + stat_3logit()
```

or standard graphics:
```{r}
plot(field0)
```



## How to specify the covariate changes

Ternary plots represent the effect of a change in covariate values on the
probability distribution of the dependent variable. The function `field3logit`
permits such change to be specified in two different ways, which are
illustrated below.

As an example, the following subsections refer to this dataset:
```{r}
data(cross_1year)
str(cross_1year)
head(cross_1year)
```

and this trinomial logistic regression model:
```{r}
mod0 <- nnet::multinom(employment_sit ~ finalgrade + irregularity + hsscore, cross_1year)
mod0
```




### Explicit specification of $\Delta x$

This method for specifying the change in the covariate values requires the
vector $\Delta x$ to be explicitly defined, thus it may be suitable when
$\Delta x$ results from some calculations. On the other hand, it is less
user-friendly than implicit syntax, as it depends on the order of regressors
in the design matrix.



#### First example

If the effect of a *high final grade* has to be assessed, the vector of changes
$\Delta x$ can be set according to the position of the dummy variable
`finalgradeHigh` in the matrix of coefficients of the model `mod0`:
```{r}
coef(mod0)
```
in this case, we have that
\[
\Delta x=[0, 0, 1, 0, 0, 0]'
\]
since `finalgradeHigh` is the fourth coefficient (including the intercept) of
the matrix of coefficients.

It follows that the function `field3logit` can be invoked as it follows:
```{r}
field0 <- field3logit(mod0, c(0, 0, 1, 0, 0, 0))
field0
```



#### Second example

It is also possible to set $\Delta x$ so as to consider changes involving more
than one regressor, as well as fractional changes. In such cases, $\Delta x$
will consist in a vector where there are several non-zero elements which may 
take any positive or negative value.

Assume, for example, that we want to study the effect of a decrease by 10 in the
high school final score, associated to an high final grade. In such a case,
we have that:
\[
\Delta x =[0, 0, 1, 0, 0, -10]'\,,
\]
thus:
```{r}
field0 <- field3logit(mod0, c(0, 0, 1, 0, 0, -10))
field0
```



### Implicit specification of $\Delta x$

Unlike the explicit method, this syntax allows the user to initialise the vector
$\Delta x$ by writing the name of the covariate which should vary: the function
`field3logit` will build $\Delta x$ up associated to a unitary change in the
specified covariate.

If more complex changes in covariate values have to be considered, implicit
syntax allows the user to express them in terms of `R` expressions involving
the covariates.



#### First example

If the effect of a *high final grade* has to be assessed, the implicit syntax
which allow to assess the effect of a unitary change of `finalgradeHigh` is the
following:
```{r}
field0 <- field3logit(mod0, 'finalgradeHigh')
field0
```

Note that the console output produced by printing `field0` shows both the
implicit effect (line `Effect`) and the associated vector $\Delta x$ (line
`Explicit effect`).




#### Second example

If we want to study the effect of a decrease by 10 in the high school final
score, associated to an high final grade, the implicit syntax is:
```{r}
field0 <- field3logit(mod0, 'finalgradeHigh - 10 * hsscore')
field0
```

Compare the line `Explicit effect` of this output to the line `Effect` of the
same example in the previous section: as expected, they are the same.






## Comparing multiple effects

When effects of multiple changes have to be compared at a time, multiple fields
should be computed and represented on the same plot. This task can be easily
done by creating a `multifield3logit` object and directly representing it.

Since objects `multifield3logit` result by putting together two or more
`field3logit` objects, the package `plot3logit` allows the user to create a
`multifield3logit` object by adding up two or more `filed3logit` or
`multifield3logit` objects using standard sum operator `+`.

Here it is an example. The following command fit a trilogit model where all
available variables are used as regressors. Then four `fields3logit` objects are
computed for assessing the effects of a some combined changes in the duration of
studies and in students' final degree score.

Note that each field is computed just with respect to a single probability
distribution (`refpoint`) of the dependent variable, and only one arrow is
computed. The reason of this is that we have to represent four fields on the
same plot, thus olny a small number of arrows can be drawn in order to preserve
the clarity of the graph.

```{r, results='hide'}
data(cross_1year)
mod0 <- nnet::multinom(employment_sit ~ ., data = cross_1year)

refpoint <- list(c(0.7, 0.15, 0.15))

field_Sdur <- field3logit(mod0, 'durationShort', label = 'Short duration', p0 = refpoint, narrows = 1)
field_Ldur <- field3logit(mod0, 'durationLong', label = 'Long duration', p0 = refpoint, narrows = 1)
field_Hfgr <- field3logit(mod0, 'finalgradeHigh', label = 'High final grade', p0 = refpoint, narrows = 1)
field_Lfgr <- field3logit(mod0, 'finalgradeLow', label = 'Low final grade', p0 = refpoint, narrows = 1)
```

Now the `multifield3logit` object can be created by adding all the `field3logit`
objects up together:
```{r}
mfields <- field_Sdur + field_Ldur + field_Lfgr + field_Hfgr
mfields
```

and the `multifield3logit` object `mfield` can be represented in a graph:
```{r, message = FALSE}
gg3logit(mfields) +
  stat_3logit(aes(colour = label)) +
  theme_zoom_L(0.45)
```




## References

Hamilton N. E., M. Ferry (2018) "ggtern: Ternary Diagrams Using ggplot2",
*Journal of Statistical Software, Code Snippets*, **87**(3), pp. 1-17.
https://doi.org/10.18637/jss.v087.c03

Santi F., M. M. Dickson, G. Espa (2019) "A graphical tool for interpreting
regression coefficients of trinomial logit models", *The American
Statistician*, **73**(2), pp. 200-207.
https://doi.org/10.1080/00031305.2018.1442368

Smith M. R. (2017). "Ternary: An R Package for Creating Ternary Plots",
*Zenodo*. https://doi.org/10.5281/zenodo.1068996

Wickham, H. (2016) *ggplot2: Elegant Graphics for Data Analysis.*
Springer-Verlag, New York.

